What We Eval in the Shadows

### 2021-04-10 paper/main.tex
Add a few more lines on why it is hard

The tracer registers callbacks to all \eval functions as well as a few other
ones that allow us to better identify where do the expressions come from. For
example, we taint the results of calls to \c{parse} (turns string into
expression) and \c{match.call} (reflects on current call). We also capture calls
to R API for dynamic code loading (\eg\xspace \c{library}, \c{require} and
\c{source}). Next, we subscribe to the events related to a variable definition
and assignment allowing us to record side effects that happen in environments
while evaluating code in \eval. One challenge we encountered during the analysis
was that essentially R only provides source references for top-level calls in a
function body or in the block surrounded by braces. That is to say, a method
whose body would \c{eval(x)} would not have debug information that we can use to
identify that particular \eval. The same expression surrounded by braces would
have a line number and source reference. This is unfortunately not easily fixed
in the R implementation. Instead, we extended the dynamic analysis tool to
attach synthetic source code references to all \eval call sites by traversing
AST. The tracer is implemented as an R package in 3.2K C++ and 1.3K of R code.
For performance reason, most of the tracing is done in C++. While in theory the
implementation is rather straightforward, what makes it hard in practice is that
(1) some of the core R functionality such as package loading is implemented
using \eval which then leaks to the tracer in unexpected places, (2) lazy
evaluation makes it harder to analyze function arguments while tracing as
prematurely forcing a promise might have consequences, and finally (3) running
large amount of real-world code triggers exercises a lot of corner cases of
highly underspecified R behavior.

>>>>

The tracer registers callbacks to all \eval functions as well as a few other
ones that allow us to better identify where do the expressions come from. For
example, we taint the results of calls to \c{parse} (turns string into
expression) and \c{match.call} (reflects on current call). We also capture calls
to R API for dynamic code loading (\eg\xspace \c{library}, \c{require} and
\c{source}). Next, we subscribe to the events related to a variable definition
and assignment allowing us to record side effects that happen in environments
while evaluating code in \eval. One challenge we encountered during the analysis
was that essentially R only provides source references for top-level calls in a
function body or in the block surrounded by braces. That is to say, a method
whose body would \c{eval(x)} would not have debug information that we can use to
identify that particular \eval. The same expression surrounded by braces would
have a line number and source reference. This is unfortunately not easily fixed
in the R implementation. Instead, we extended the dynamic analysis tool to
attach synthetic source code references to all \eval call sites by traversing
AST. The tracer is implemented as an R package in 3.2K C++ and 1.3K of R code.
For performance reason, most of the tracing is done in C++. While in theory the
implementation is rather straightforward, it is not so in practice:
%
\begin{compactitem}[---]

\item The lazy evaluation makes it difficult to analyze function arguments
  while tracing as prematurely forcing a promise might have dangerous
  consequences.

\item While the R interpreter is implemented in C, a lot of the core
  functionality is done in R. For example, package loading is implemented in R
  using \eval or the S4 and R6 object systems. This makes it hard in the tracer
  to separate the \eval that are essential to user programs from the
  accidental ones that are products of the way R implements its basic
  operations. This is even more so for the side-effects analysis.

\item In the dynamic analysis, we run all the \emph{runnable} code we obtained
  from CRAN and Kaggle---\ie real code written by a people with highly varying
  expertise in R and programming in general. This exercise a lot of the corner
  cases of the highly underspecified R behavior.

\end{compactitem}
